---
title: "How to in R"
subtitle: "datasetjson - Read and write CDISC Dataset JSON formatted datasets in R and Python"
author: <code>R/Pharma 2025 Workshop</code>
date: 2025-11-07
logo: "../../images/sticker.svg"
footer: "[https://github.com/atorus-research/datasetjson_workshop](https://github.com/atorus-research/datasetjson_workshop)" 
editor: source
engine: knitr
format: 
  live-revealjs: 
    theme: ../slides.scss
    transition: fade
    slide-number: true
    chalkboard: true
filters:
  - output-line-highlight.lua
execute:
  echo: true
  freeze: false
cache: false
---

{{< include ../../_extensions/r-wasm/live/_knitr.qmd >}}

## The Whole Game

Let's see the complete workflow with [{pharmaverseadam}](https://github.com/pharmaverse/pharmaverseadam) data:

::: incremental
1.  **Start** with ADAM datasets (RDS/Parquet/xpt/sas7bdat/...)
2.  **Prepare** top-level metadata and column metadata
3.  **Convert** to Dataset-JSON
4.  **Share** the standardized file
5.  **Read** it back anywhere
:::

------------------------------------------------------------------------

## Step 1: Load ADAM Data and Metadata

```{r}
#| eval: true
#| code-line-numbers: 1-2|3-5
library(datasetjson)

# Load data (RDS, Parquet, ... - your choice)
adsl <- readRDS("../../data/adam/adsl.rds")
# adsl <- arrow::read_parquet("adsl.parquet")
```

------------------------------------------------------------------------

## Step 1a: Examine What We Have - ADSL

```{r}
#| eval: true
knitr::kable(head(adsl[1:6]), format = "html", table.attr = "style='font-size: 24px;'")
```

------------------------------------------------------------------------

## Step 2: Prepare Metadata

```{r}
#| eval: true
#| 
# Load column metadata
adsl_meta <- readRDS("../../data/adam/metadata/adsl_meta.rds")
# adsl_meta <- arrow::read_parquet("adsl_meta.parquet")
```

------------------------------------------------------------------------

## Step 2a: Examine What We Have - adsl_meta

```{r}
#| eval: true
knitr::kable(head(adsl_meta), format = "html", table.attr = "style='font-size: 24px;'")
```

<!-- --- -->

<!-- ## Step 2: Examine What We Have -->

<!-- #### Data Type: float -->

<!-- ```{r} -->

<!-- #| eval: true -->

<!-- floats <- adsl_meta |> -->

<!--   dplyr::filter(adsl_meta$dataType == "float") -->

<!-- knitr::kable( -->

<!--   head(floats, 4), -->

<!--   format = "html", -->

<!--   table.attr = "style='font-size: 24px;'") -->

<!-- ``` -->

<!-- --- -->

<!-- ## Step 2: Examine What We Have -->

<!-- #### Data Type: datetime -->

<!-- ```{r} -->

<!-- #| eval: true -->

<!-- datetime <- adsl_meta |> -->

<!--   dplyr::filter(adsl_meta$dataType == "datetime") -->

<!-- knitr::kable( -->

<!--   head(datetime, 4), -->

<!--   format = "html", -->

<!--   table.attr = "style='font-size: 24px;'") -->

<!-- ``` -->

<!-- --- -->

<!-- ## Step 2: Examine What We Have -->

<!-- #### Data Type: date -->

<!-- ```{r} -->

<!-- #| eval: true -->

<!-- dates <- adsl_meta |> -->

<!--   dplyr::filter(adsl_meta$dataType == "date") -->

<!-- knitr::kable( -->

<!--   head(dates, 4), -->

<!--   format = "html", -->

<!--   table.attr = "style='font-size: 24px;'") -->

<!-- ``` -->

------------------------------------------------------------------------

## Step 3: Create Dataset-JSON

```{r}
#| eval: true
#| code-line-numbers: 1,5,10|1,2,7,8|1,3,9|1,4,6
# Convert ADSL to Dataset-JSON
## top-level metadata
## column/variable metadata
## rows
adsl_json <- dataset_json(
  adsl,
  name = "ADSL", 
  dataset_label = "Subject Level Analysis Dataset",
  columns = adsl_meta
)
```

::: aside
<a href="https://github.com/cdisc-org/DataExchange-DatasetJson/blob/40b2b23c6792d073bed1ee32e15891077ea90d96/doc/dataset-json1-1.md" target="_blank">CDISC Dataset-JSON v1.1 Specification</a>
:::

------------------------------------------------------------------------

## Step 3a: Inspect Dataset-JSON Object

```{r}
#| eval: true
head(adsl_json)
```

------------------------------------------------------------------------

## Step 3b: Inspect Dataset-JSON Object Attributes

#### class, top-level, column

```{r}
#| eval: true
#| class-output: highlight
#| output-line-numbers: 5|4,7,8|9-1000
str(attributes(adsl_json))
```

------------------------------------------------------------------------

## Step 4: Share Standardized Files

```{r}
#| eval: true
# Write to Dataset-JSON files
write_dataset_json(adsl_json, "output/ADSL.json", pretty = TRUE) # use pretty when viewing
```

<br><br> Now anyone can read these files!

-   Regulatory reviewers
-   External collaborators
-   Different software (R, Python, SAS)

------------------------------------------------------------------------

## Step 4a: Inspect Dataset-JSON File

```{ojs}
FileAttachment("output/ADSL.json").json()
```

------------------------------------------------------------------------

## Step 5: Read Back Anywhere

```{r}
#| eval: true
# Read back - gets original data + metadata
adsl_restored <- read_dataset_json("output/ADSL.json")
```

------------------------------------------------------------------------

## Step 5a: Inspect Dataset-JSON Object

```{r}
#| eval: true
adsl_restored
```

------------------------------------------------------------------------

## Step 5b: Inspect Dataset-JSON Object Attributes

#### class, top-level, column

```{r}
#| eval: true
#| class-output: highlight
#| output-line-numbers: 4|5,6,7|7-1000
str(attributes(adsl_restored))
```

<!-- ------------------------------------------------------------------------ -->

<!-- ## Step 6: Read Back Anywhere -->

<!-- ```{r} -->

<!-- #| eval: true -->

<!-- # Read back - gets original data + metadata -->

<!-- str(attributes(adsl_restored)[c("label", "name", "columns")]) -->

<!-- ``` -->

------------------------------------------------------------------------

## Compare Original and Restored

```{r}
#| eval: true
# Identical to original
diffdf::diffdf(adsl, adsl_restored)
```

------------------------------------------------------------------------

## Compare Original and Restored

```{r}
#| eval: true
# compare class and attributes
waldo::compare(adsl, adsl_restored)
```

------------------------------------------------------------------------

## Why This Matters

::: incremental
-   **Format agnostic**: Start with RDS, Parquet, CSV - doesn't matter, convert to RDS
-   **Metadata preserved**: Labels, types, all CDISC information
-   **Standardized**: One format for sharing across tools/organizations
-   **Validated**: Built-in schema validation
-   **Round-trip safe**: *Almost* Perfect data fidelity
:::

------------------------------------------------------------------------

## *Almost!?!* Perfect Data Fidelity

::: incremental
-   [R doesn’t have a specific built in type of time. We decided to take on {hms} as a dependency.](https://atorus-research.github.io/datasetjson/articles/date_time_datetime.html)
    -   Takeaway: You can see a difference in classes and formatting between writing and reading times. We will always apply \`hms::hms()\` to times on read.
-   [Numeric precision starts to vary around the 7th decimal](https://atorus-research.github.io/datasetjson/articles/precision.html)
    -   Takeaway: Set the `float_as_decimals` and `decimals_as_floats` arguments to TRUE when writing and reading to preserve 16 digits
:::

------------------------------------------------------------------------

## Time Demo

```{r}
#| eval: true
#| code-line-numbers: 1-6|7-22|23-33|34-40
# Load libraries and data
library(datasetjson)
library(dplyr)

time_df <- head(iris, 5)

# add a time
time_df['time_col'] <- lubridate::hms(
  c("10:00:00", "11:30:15", "14:45:30", "18:01:00", "22:15:59")
  )

# add column metadata for this new time var
time_items <- iris_items |> bind_rows(
  data.frame(
    itemOID = "IT.IR.float_col",
    name = "time_col",
    label = "Test time class",
    dataType = "time",
    targetDataType = "integer"
  )
)

# create datasetjson object and write out
dsjson <- dataset_json(
  time_df, 
  item_oid = "time_df",
  name = "time_df",
  dataset_label = "time_df",
  columns = time_items
)

json_out <- write_dataset_json(dsjson)

# read back in
out <- read_dataset_json(json_out)
```

------------------------------------------------------------------------

## Time Demo - Show class difference

```{r}
#| class-output: highlight
#| output-line-numbers: 1-5|6-7
# class compare
classes <- list(original = class(time_df$time_col) , new =  class(out$time_col))
classes
```
------------------------------------------------------------------------

## Time Demo - Values

```{r}
#| class-output: highlight
#| output-line-numbers: 1|2|3|4|5
# value compare
for (i in 1:length(time_df$time_col)) {
  print(paste("Value from original:", time_df$time_col[i], "| Value from new:", out$time_col[i]))
}
```

------------------------------------------------------------------------

## Numeric Precision Demo
#### Show decimal precision difference

```{r}
#| eval: true
#| code-line-numbers: 1-6|7-15|16-25|26-36|37-40
# Load libraries and data
library(datasetjson)
library(dplyr)

test_df <- head(iris, 5)

# add a float/double
test_df['float_col'] <- c(
  143.66666666666699825,
  2/3,
  1/3,
  165/37,
  6/7
)

# add column metadata for this new float/double
test_items <- iris_items |> bind_rows(
  data.frame(
    itemOID = "IT.IR.float_col",
    name = "float_col",
    label = "Test column long decimal",
    dataType = "float"
  )
)

# create datasetjson object and write out
dsjson <- dataset_json(
  test_df, 
  item_oid = "test_df",
  name = "test_df",
  dataset_label = "test_df",
  columns = test_items
)

json_out <- write_dataset_json(dsjson)

# read back in and subtract
out <- read_dataset_json(json_out)

test_df$float_col - out$float_col
```

## Numeric Precision Demo
#### Use `float_as_decimals` and `decimals_as_floats` arguments

```{r}
json_out <-write_dataset_json(dsjson, float_as_decimals = TRUE)

out <- read_dataset_json(json_out, decimals_as_floats = TRUE)

test_df$float_col - out$float_col
```


------------------------------------------------------------------------

## Your Turn: Exercises

Now repeat the whole game with ADAE data! Open `exercises/01-r.R`:

::: incremental
1.  **Load ADAE** - Load data and metadata (RDS or Parquet)
2.  **Examine** - Look at data types and structure
3.  **Convert** - Create Dataset-JSON object
4.  **Write** - Save to standardized file
5.  **Read Back** - Verify round-trip fidelity
:::

## Exercise Preview

```{r}
#| eval: false
# Load ADAE data and metadata
adae <- readRDS("data/adam/adae.rds")
adae_meta <- readRDS("data/adam/metadata/adae_meta.rds")

# Follow the same steps as ADSL!
# Convert, write, read back, and compare
```

## Questions?

::: center
Ready to dive into the exercises?

Open `exercises/01-r.R` and let's explore the datasetjson package together!
:::
